---
title: "Data processing"
output: html_notebook
---

Ctrl + Shift + Enter - run cell.
Ctrl + Alt + I - insert new cell.
Ctrl + Shift + K - preview HTML result document.

# Download data

```{r}
file_path = './Data/OnlineNewsPopularity.csv'
data = read.csv(file_path)
```


```{r}
dim(data)
head(data)
```



## When the majority of the publications are written?

Questions to answer:
1. When the majority of publications are written
2. Check week_day_is_* parameter - may be articles, written on a certain day, had
the highest number of shares?

```{r}
par(mfrow = c(2, 2))

hist(data$weekday_is_friday)
hist(data$weekday_is_monday)
hist(data$weekday_is_saturday)
hist(data$weekday_is_sunday)

par(mfrow = c(1, 1))
```
```{r}
find_majority <- function(data_, columns){
  sizes = c()
  colnames = c()
  for(c in columns){
    size = dim(data_[data_[, c] == T, ])[1]
    sizes = append(sizes, size)
    colnames = append(colnames, c)
    #res = append(res, list('size' = size, 'column' = c))
  }
  # print(sizes)
  # print(which.max(sizes))
  cat('Maximum of articles is published on:', colnames[which.max(sizes)], max(sizes), 'articles', '\n')
  return(sizes)
}

articles_amount = find_majority(
  data, 
  columns = c(
    'weekday_is_monday', 
    'weekday_is_tuesday',
    'weekday_is_wednesday',
    'weekday_is_thursday',
    'weekday_is_friday',
    'weekday_is_saturday',
    'weekday_is_sunday'
  )
)
```
## Is there a difference in a number of shares, depending on the day of publication?

Here, we will use statistical tests.
For comparing number of shares on weekdays and weekends -> we will use t-test.
What are assumptions of t test?
1. Number of shares of one article is independent of number of shares of the other article (debatable)
2. Shares are normally distributed (not true)

```{r}

hist(data$shares)
max(data$shares)
min(data$shares)
```



```{r}
vec = seq(1:dim(data)[1])# vector(length = dim(data)[1])

vec[which(data$weekday_is_monday == T)] = 'mon'
vec[which(data$weekday_is_tuesday == T)] = 'tue'
vec[which(data$weekday_is_wednesday == T)] = 'wed'
vec[which(data$weekday_is_thursday == T)] = 'thur'
vec[which(data$weekday_is_friday == T)] = 'fri'
vec[which(data$weekday_is_saturday == T)] = 'sat'
vec[which(data$weekday_is_sunday == T)] = 'sun'

vec_factor = factor(
  vec, 
  levels = c('mon', 'tue', 'wed', 'thur', 'fri', 'sat', 'sun', labels = levels)
)

data$weekday = vec_factor
data$is_weekend = data$weekday_is_sunday | data$weekday_is_saturday

```


```{r}
par(mfrow = c(2, 2))

boxplot(data$shares ~ data$is_weekend, outline = F)
boxplot(data$shares ~ data$weekday, outline = F)

boxplot(log(data$shares) ~ data$is_weekend, outline = F)
boxplot(log(data$shares) ~ data$weekday, outline = F)

par(mfrow = c (1, 1))
```


```{r}

```






